The galactic chemical evolution of carbon: Implications for stellar nucleosynthesis

### 2025-11-14 carbon_paper/tex/main.tex
near-submission draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,
referee, % this line changes to double-spaced 1 column
usenatbib]{mnras}

@@

@@

\newcommand{\fruity}{\texttt{\hyperlink{fruity}{Fruity}}}
\newcommand{\nugrid}{\texttt{\hyperlink{nugrid}{NuGrid}}}
\newcommand{\monash}{\texttt{\hyperlink{monash}{Monash}}}
\newcommand{\aton}{\texttt{\hyperlink{aton}{ATON}}}
\newcommand{\cfactor}{2.4}
\newcommand{\zetaobest}{1.90 \pm 0.10}
\newcommand{\zetaibest}{2.28 \pm 0.16}
\newcommand{\zetaiibest}{2.59 \pm 0.30}

@@

% internal abbreviations
\newcommand{\caah}{[C/Mg]-[Mg/H]}
\newcommand{\caafe}{[C/Mg]-[Mg/Fe]}

@@

@@

\newcommand{\Yct}{{y_{\rm C}}}
\newcommand{\Ycc}{{y_{\rm C}^{\rm CC}}}
\newcommand{\Yoc}{{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{{y_{\rm C}^{\rm AGB}}}
\newcommand{\ycagb}{\y_{\rm C}^{\rm AGB}}
\newcommand{\aagb}{\alpha_{\rm C}^{\rm AGB}}
\newcommand{\zagb}{\zeta_{\rm C}^{\rm AGB}}
\newcommand{\zcc}{\zeta_{\rm C}^{\rm CC}}
\newcommand{\noneqagb}{\nu_{\rm C}^{\rm AGB}}
 \newcommand{\yb}{\ensuremath{\rotatebox[origin=B,y=0.5ex]{180}{y}}}
\newcommand{\y}{p}
\newcommand{\fagb}{f_{\rm C}^{\rm AGB}}

\newcommand{\zetao}{\zeta_0}
\newcommand{\zetai}{\zeta_{1}}
\newcommand{\zetaii}{\zeta_{2}}
\newcommand{\Zl}{x_{\rm low}}

\newcommand{\Mo}{ {\rm M}_{\sun}}

\newcommand{\Zo}{ Z_{\sun}}

@@

%%%%%%%%%%% JWJ' editing macros %%%%%%%%%%%
\newcommand{\strike}[1]{{\color{ForestGreen} \sout{#1}}}
\newcommand{\add}[1]{{\color{ForestGreen} #1}}
\newcommand{\note}[1]{{\color{ForestGreen} \textit{ \small (JWJ: #1)}}}
% \LetLtxMacro\origcitep\citep
% \LetLtxMacro\origcitet\citet
% \renewcommand{\citep}[2][]{%
%     \mbox{\origcitep[#1][]{#2}}
% }
% \renewcommand{\citet}[2][]{%
%     \mbox{\origcitet[#1][]{#2}}
% }

@@

% \LetLtxMacro\origcite\cite

% This is my command. I want to strike out, 
% \newcommand{\strikeit}[1]{%
%   \ifcorrectingmode
%   \mbox{\sout{#1}}%
% %  \makebox[\textwidth][s]{\sout{#1}}%
%   \fi
% }

% \renewcommand{\cite}[2][]{%
%   \ifcorrectingmode
%   \mbox{\origcite[#1]{#2}}%
%   \else
%   \origcite[#1]{#2}%
%   \fi
% }

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[The Origin and Galactic Evolution of Carbon]{The Galactic Chemical Evolution of Carbon: Implications for Stellar Nucleosynthesis }

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[D. A. Boyea et. al.]{%
Daniel A. Boyea,$^{1, 2, 3}$\thanks{%
Contact e-mail:~\href{mailto:boyea.2@osu.edu}{boyea.2@osu.edu}}
James W. Johnson,$^{1, 2, 4}$
Third Author$^{2,3}$
and Others$^{1,3}$
\\
% List of institutions
$^{1}$Department of Astronomy, The Ohio State University, 140 W. 18th Ave., Columbus, OH, 43210, USA
\\
$^{2}$Center for Cosmology \& Astroparticle Physics (CCAPP), The Ohio State University, 191 W. Woodruff Ave., Columbus, OH, 43210, USA
\\
$^{3}$Department of Physics \& Astronomy, University of Victoria, Victoria, BC V8P 5C2, Canada
\\
$^{4}$The Observatories of the Carnegie Institution for Science, 813 Santa Barbara St., Pasadena, CA, 91101, USA
}

@@

% Abstract of the paper
\begin{abstract}
% context
The astrophysical origin of carbon remains largely uncertain---different studies propose that either asymptotic giant branch (AGB) stars or core collapse supernovae (CCSNe) may be the dominant process.
% 
We aim to constrain the stellar yields of C through multi-zone Galactic chemical evolution models by comparing predictions with \apogee\ subgiants abundances.
% 
We show \caah\ trend represents the equilibrium abundances of C and Mg, deriving the total C yield as a function of metallicity. 
Instead, the \caafe\ trend (at a given metallicity) is an empirical estimate of the delay time distribution of C.
We infer that AGB stars produce between 10 and 40\% of C. Additionally, we show that all AGB models produce C too quickly, and the low-mass AGB C yield is likely underestimated. 
% misc
Our models are relatively independent of uniform scaling of yields and outflows, and alternate star formation histories. 
However, the stars which contribute to \agb\ C production and the \ia\ delay time distribution of Fe contribute uncertainties to our conclusions. 
% gas phase
While reliable gas-phase and low-metallicity measurements of C are challenging, we find that our model and a single-zone model with our recommended yields replicate the broad trends of \caah{} across different environments and metallicities.

\end{abstract}

@@

\begin{figure*}
    \centering
    \includegraphics{subgiants.pdf}
    \caption{The [C/Mg] ratio against [Mg/H] (left) and [Mg/Fe] (right) for the \citet{jack}~sample of \apogee{} subgiants. On the left, we plot high and low-$\alpha$ stars in blue and orange, using the separation defined in Eq.~\ref{eq:high_alpha}.  On the right, we colour-code stars according to their [Mg/H] abundance. The median subgiant trends serve as our emperical benchmark. } \label{fig:subgiants}
\end{figure*}

Despite being the most abundant metal in the Universe~\citep[e.g.][]{magg+22}, the nucleosynthetic origin of C is poorly understood. 
There is broad agreement that both high mass and asymptotic giant branch (AGB) stars produce C \citep{jennifer19}, but their relative yields are disputed.
From a Galactic chemical evolution (GCE) perspective, some authors have argued that high-mass stars should dominate~\citep[e.g.][]{prantzos+94, HEK00, romano+20, franchini+20, gustafsson22}, and others have argued that AGB stars should contribute at least half of the C in the universe~\citep[e.g.][]{tinsley79, chiappini+03, mattsson10, KKU11, rybizki+17, KKL20}. 
In this paper, we aim to constrain C production using subgiant stars and multi-zone GCE models.% \dbstrike{incorporating stellar radial migration}.

% Formed in the cores of stars during He fusion, C is the first \dbadd{primary? element} after He. Additionally, C is one of the only light elements formed in low-mass stars \citep[e.g.,][]{KL14}. \dbnote{I don't like how many e.g.'s there are, and is this the right citation here?}
%\dbadd{
% Because the effects of first dredge up are mass-dependent, C and N abundances are frequently used as age indicators for RGB stars \citep{MG15, martig16, hasselquist19, vincenzo+21}. 
% }
% \dbnote{C abundnces for stellar structure (opacities) and star formation (line cooling)}
% \dbadd{In addition, stellar parameters such as convection and mass loss are calibrated to match observed stars and cannot be predicted from first principles.} \dbnote{need to check if this means SE models are calibrated to match yields or just stellar parameters.}

Predictions of C yields from stellar evolution and supernova (SN) models are riddled with uncertainties (see, e.g., the reviews by \citealt{romano+10, KL14}).
Some salient unknowns include mass loss rates \citep{sukhbold+16, beasor+2020}, rotational mixing \citep{frischknecht+16, LC18}, nuclear reaction rates \citep{herwig+05, herwig+austin2004}, and convection \citep{chieffi2001, ventura+13}.
Recent work additionally suggests that small changes in a stars's initial mass of 
$\sim$$0.01 \Mo$  have significant effects on the evolution and explosion outcomes of massive stars \citep{bruenn+2023, vartanyan_burrows2023}.
 Furthermore, many stars are found in binary systems \citep{sana+12}, the effects of which are almost entirely
unexplored in stellar nucleosynthesis models \citet{farmer+21}.
Accurate predictions of metal yields from stellar evolution models remain elusive.

Given the challenge of theoretical yield predictions, recent work aims to empirically characterize elemental production. \dbnote{are there older relevant references like garnett1997? or other relevant work (eitner+20?)}
\citet{weinberg+19, weinberg+22} analyzed the abundance ratio trends of high- and low-alpha sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and their metallicity dependence.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}.
\citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSNe) using the radioactive tails of their lightcurves.
Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.

 % Folowing many previous works  (e.g. \citealt{DTS78, BF06, prantzos+18, berg+19}; see also the recent reviews by \citealt{romano22} and \citealt{RM21}.) this paper focuses on the evolution of C in GCE models.
\citet{james+24} recently argued that the enrichment history of the Galactic disk can be described as an equilibrium phenomenon.
In equilibrium GCE, trends in abundance ratios closely trace yield ratios (see their section 5.2.2).
\citet{james+23} had previously shown that the correlation between gas-phase N and O abundances (see their fig. 1) could be described as a consequence of the metallicity dependence of the relative N and O yields.
Under their equilibrium scenario, this argument should extend to any element whose enrichment by stellar populations does not require long timescales ($\gtrsim$few Gyr).
We show in section~\ref{sec:agb} that the characteristic delay times of C production should be $\sim$ 1 Gyr, within this limit. 
\dbnote{This is complicated by the difference between high alpha and low alpha, and 1-2Msun stars do produce carbon.... Also, I would like to call out reviews and general works on C at some point.}

We use a sample of subgiant stars from APOGEE \citep{apogee17} as our primary observational constraint.
APOGEE Subgiants provide a large, homogeneous sample of well-measured C abundances unaffected by stellar evolution.
\dbadd{Subgiants' photospheric C abundance closely matches their birth composition.}
\dbstrike{Subgiants have the advantage that their photospheric abundances of C directly reflect their birth C abundances much more closely than other spectral types.} \citep{gilroy89, korn+07, lind+08, souto+18, souto19}.
In other evolutionary phases, the measured abundances are known to be affected by internal processes (see discussion in section~\ref{sec:data_selection} below).
Our sample should therefore be unaffected by uncertainties with inferring their birth abundances.

In section~\ref{sec:data_selection} we describe the selection for our subgiant observational sample.
In section~\ref{sec:nucleosynthesis}, we discusses theoretical estimates of \agb\ and \cc\ C yields and present the yield prescriptions we use for our models.
In section~\ref{sec:vice}, we describe the remaining details of our GCE models.
In section~\ref{sec:results}, we present our model predictions and discuss their observationally distinguishable predictions. 
In section~\ref{sec:gas}, we compare our model to observations of halo stars and extragalactic HII regions from the literature. 
We summarize our conclusins in section~\ref{sec:conclusions}.

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (\textit{gravitational settling}; e.g. \citealt{turncotte+98}).
When stars evolve off main sequence and become subgiants, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
However, once a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with core, CNO-processed material
\citep{iben67, KL14}. 
\dbnote{Is there other justification for subgiants? Consistent stellar parameters, subgiant spectra are easier than white dwarfs? only a $\sim 0.05$ dex correction}

We use the sample of \apogee\ DR17 subgiants from \citet{jack}. We select stars from the ASPCAP pipeline (revision 1) within the $\log g$-$T_\text{eff}$ polygon given by
 \begin{equation}
    \begin{cases} \label{eq:subgiant_selection}
        \log \text{g} \geq 3.5 \\
        \log \text{g} \leq 0.004\,T_{\rm eff} - 15.7 \\
        \log \text{g} \leq 0.0007\,T_{\rm eff} + 0.36 \\
        \log \text{g} \leq -0.0015\,T_{\rm eff} + 12.05 \\
        \log \text{g} \geq 0.0012\,T_{\rm eff} - 2.8. \\
    \end{cases}
\end{equation}
Following \citet{jack}, we also exclude stars with the flags:
        \verb|APOGEE_MIRCLUSTER_STAR|,
        \verb|APOGEE_EMISSION_STAR|,
        \verb|APOGEE_EMBEDDEDCLUSTER_STAR|,
        \verb|APOGEE2_YOUNG_CLUSTER|,
        \verb|APOGEE2_W345|,
        \verb|APOGEE2_EB|, and
        \verb|DUPLICATE|.
We additionally remove stars with no reported (or flagged unreliable), C, N, Mg, or Fe abundances. These cuts yield a sample containing \nsubgiants\ subgiants.

Fig.~\ref{fig:subgiants} shows out subgiant sample plotted in the \caah\ and \caafe{} planes.\footnotemark{} In this sample, [C/Mg] increases with [Mg/H], and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H]. Our aim is to reproduce these trends.
Low-$\alpha$ stars are defined to have 
\begin{equation}\label{eq:high_alpha}
\begin{cases}
\text{[Mg/Fe]} <0.16-0.13\,\text{[Fe/H]}, & \text{[Fe/H]}<0\\
\text{[Mg/Fe]} <0.16, & \text{[Fe/H]}>0. \\
\end{cases}
\end{equation}

\footnotetext{In this paper, we use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the mass ratio between A and B, scaled such that $[A/B]=0$ for the sun (see Table~\ref{tab:fiducial_mod} for solar scale.) }

We have also investigated these thrends in other catlogues (not show) in order to understand the systematic uncertainties of our APOGEE subgiant sample.
Namely, we we consider subgiants/dwarfs from GALAH DR4 \citep{galah_dr4}, Gaia-ESO DR5.1 \citep{gaiaeso_dr5.1}, and LAMOST \citep{lamost} with similar selection criteria. Additionally, we consider APOGEE Giants with first dredge up mixing corrections from \citep{vincenzo+21}.
In general, different surveys follow similar trends in \caah\ and \caafe\ but with systematic offsets at the $\sim{0.1}$ dex level in [C/Mg] given [Mg/H] and slight variations in the slope.
These systematic differences are not a major concern for our conclusions since our models are mostly constrained by the slope of \caah\ and \caafe,  not their normalization (see Fig.~\ref{fig:zeta_f} and discussion in section \dbadd{5.2}).

\section{Nucleosynthesis}\label{sec:nucleosynthesis}

A yield quantifies the new production of a chemical species. For a given star, we define the {\it stellar yield of $X$} as the fraction of a star's initial mass synthesized into a new element, $X$. Specifically, the stellar yield of $X$ is
\begin{equation} \label{eq:stellar_yield_def}
   \y_{\rm X}(m, Z) := 
    \frac{m_{X, \rm ej}- Z_{X,\rm  ini}\,m_{\rm ej}}{m}
\end{equation}
where $Z_{X,\rm ini}$ is the initial mass abundance of $X$, $m_{\rm ej}$ is the ejected mass, $m_{X, \rm ej}$ is the ejected mass of $X$, and $m$ is the initial mass of the star.
Note that yields may become negative if more X is transformed into other elements than is produced.
We also describe yields by their population-averaged quantities, {\it integrated yields}, the amount of newly produced X per unit mass of star formation. 
The integrated yield of $X$ is
\begin{equation} 
   y_{X}(Z) := 
    \frac{M_X - r\,Z_{X, \rm ini}\,M_{\rm ssp}}{M_{\rm ssp}},
\end{equation}
where $M_X$ is the total expelled mass of element X, $M_{\rm ssp}$ is the birth mass of the single stellar population (SSP)\footnote{An SSP is a cluster of stars born at the same time from the same material.}, and $r$ is the return fraction (fraction of the SSP mass returned to the interstellar medium).

Table~\ref{tab:fiducial_mod} summarizes our yield choices. 
Following \citet{james+23}, we integrate yields and our GCE models using the publicly available Versatile Integrator for Chemical Evolution (\VICE, \citealt{JW20}). Our solar abundance scale is from \citet{magg+22} with a 0.04 dex gravitational settling correction. Also from \citet{magg+22}, we adopt $\Zo = 0.0176$. We use the O, Mg, and Fe yields recommended by \citet{david_fe} based on measurements of the mean CCSN Fe yield \citep{rodriguez+23} and APOGEE chemical abundances.
Our N yield is scaled from \citep{james+23}.
We assume a \citet{kroupa01} initial mass function (IMF).

\begin{table}
	\centering
    \caption[]{Solar abundance scale and fiducial yields (in units of stellar population birth mass). See section \ref{sec:agb} for the definition of \fruity. The solar abundance scale is \citet{magg+22} + 0.04 dex suggested by \citet{david_fe}.
        \dbnote{Technically, I believe this abundance scale is off by the hydrogen 1 atomic weight. But this is < the systematic correction so I think this is negligable?}.
    \dbnote{any tips to make this table a little bit smaller?}
    }
	\label{tab:fiducial_mod}

	\begin{tabular}{l l l l l}
		\hline
         & $Z_{X,\,\sun}$ & $y_X^{\rm cc}$ & $\y_X^{\rm agb}$ & $y_X^{\rm ia}$  \\
		\hline
        C & $33.9\times10^{-4}$ & Eq.~\ref{eq:y_cc} & $\cfactor\times$\fruity &  0 \\
        O & $73.3\times10^{-4}$ & $71.3\times10^{-4}$ & 0 & 0 \\
        Mg & $6.71\times10^{-4}$ & $6.52\times 10^{-4}$ & 0 & 0 \\
        Fe & $13.7\times10^{-4}$ & $4.73\times10^{-4}$ & 0 & $7.82\times10^{-4}$ \\
        N & $10.4\times10^{-4}$ & $5\times10^{-4}$ & $5\times10^{-4}M\left(\frac{Z}{\Zo}\right)$ & 0\\
		\hline
	\end{tabular}
\end{table}

\subsection{Asymptotic Giant Branch Stars}\label{sec:agb}

\begin{figure*}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[]{The net fractional \agb\ C yield  plotted as a function of initial stellar mass and colour-coded according to metallicity. The black dashed line shows $\Ycagb=0$ for reference. Each panel represents yields from one of four \agb\ models: \fruity{}, \aton{}, \monash{}, \nugrid{} (see sections \ref{sec:agb}) }.

@@

\dbadd{data and codes used in this paper are all publicly available. (do i need links here?). }

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,
%onecolumn,
%doublespacing, % this line changes to double-spaced 1 column
%referee,
usenatbib]{mnras}

@@

% enable line numbers
\usepackage{lineno}
%\linenumbers

@@

\newcommand{\fruity}{\texttt{\hyperlink{fruity}{Fruity}}}
\newcommand{\nugrid}{\texttt{\hyperlink{nugrid}{NuGrid}}}
\newcommand{\monash}{\texttt{\hyperlink{monash}{Monash}}}
\newcommand{\aton}{\texttt{\hyperlink{aton}{ATON}}}
\newcommand{\cfactor}{2.57}
\newcommand{\zetaobest}{1.90 \pm 0.10}
\newcommand{\zetaibest}{2.28 \pm 0.16}

@@

% internal abbreviations
\newcommand{\caah}{[C/Mg]-[Mg/H]}
\newcommand{\caafe}{[C/Mg]-[Mg/Fe]}
\newcommand{\oh}{{\rm [O/H]}}
\newcommand{\mgh}{{\rm [Mg/H]}}
\newcommand{\cmg}{{\rm [C/Mg]}}
\newcommand{\co}{{\rm [C/O]}}
\newcommand{\mgfe}{{\rm [Mg/Fe]}}

@@

\newcommand{\Hii}{H~{\sc ii}}

@@

\newcommand{\Yct}{{y_{\rm C}^{\rm  tot}}}
\newcommand{\Ycc}{{y_{\rm C}^{\rm CC}}}
\newcommand{\Ymg}{{y_{\rm Mg}^{\rm CC}}}
\newcommand{\Ycagb}{{y_{\rm C}^{\rm AGB}}}
\newcommand{\ycagb}{\y_{\rm C}^{\rm AGB}}
\newcommand{\y}{p}

\newcommand{\aagb}{\beta_{\rm C}^{\rm AGB}}
\newcommand{\fagb}{f_{\rm C}^{\rm AGB}}
\newcommand{\yo}{y_0}
\newcommand{\yl}{y_{\rm low}}
\newcommand{\zcc}{\zeta_{\rm C}^{\rm CC}}

\newcommand{\Mo}{ {\rm M}_{\sun}}
\newcommand{\Zo}{ Z_{\sun}}

@@

%%%%%%%%%%% JWJ' editing macros %%%%%%%%%%%
\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\strike}[1]{{\color{red} \sout{#1}}}
% \newcommand{\add}[1]{{\color{ForestGreen} #1}}
% \newcommand{\note}[1]{{\color{ForestGreen} \textit{ \small (JWJ: #1)}}}
% \LetLtxMacro\origcitep\citep
% \LetLtxMacro\origcitet\citet
% \renewcommand{\citep}[2][]{%
%     \mbox{\origcitep[#1][]{#2}}
% }
% \renewcommand{\citet}[2][]{%
%     \mbox{\origcitet[#1][]{#2}}
% }

@@

@@

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[The origin and galactic evolution of carbon]{The galactic chemical evolution of carbon: Implications for stellar nucleosynthesis}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[D. A. Boyea et. al.]{%
Daniel A. Boyea,$^{1, 2, 3}$\thanks{%
Contact e-mail:~\href{mailto:danielaboyea@gmail.com}{danielaboyea@gmail.com}}
James W. Johnson,$^{1, 2, 4}$
and others
%David H. Weinberg,$^{1,2}$
%Jennifer A. Johnson,$^{1,2}$
%Falk Herwig$^{3}$
\\
% List of institutions
$^{1}$Department of Astronomy, The Ohio State University, 140 W. 18th Ave., Columbus, OH, 43210, USA
\\
$^{2}$Center for Cosmology \& Astroparticle Physics (CCAPP), The Ohio State University, 191 W. Woodruff Ave., Columbus, OH, 43210, USA
\\
$^{3}$Department of Physics \& Astronomy, University of Victoria, Victoria, BC V8P 5C2, Canada
\\
$^{4}$The Observatories of the Carnegie Institution for Science, 813 Santa Barbara St., Pasadena, CA, 91101, USA
}

@@

% Abstract of the paper
\begin{abstract}
Carbon (C) is thought to be produced by both core collapse supernovae (CCSN) and asymptotic giant branch (AGB) stars, but the relative contributions of these two sources are uncertain. 
% goals
We investigate the astrophysical origin of C using models of Galactic chemical evolution (GCE) appropriate for the Milky Way disk.
We benchmark our results against APOGEE subgiant abundances.
% results
The trend between [C/Mg] and [Mg/H] is set by the total C yield as a function of metallicity. Observations indicate a gently rising [C/Mg] with [Mg/H], but AGB C production is predicted to decline with metallicity. 
Our sample therefore favours a scenario in which CCSN yields rise with metallicity to offset declining AGB C yields and drive a subtle increase in [C/Mg] with [Mg/H].
This result is consistent with massive star nucleosynthesis models incorporating rotation.
%
The [C/Mg]-[Mg/Fe] trend is sensitive to delayed enrichment and therefore constrains the amount of AGB C production.
Given the slope of this relation, we find that AGB stars likely account for 10--40 per cent of C at solar metallicity.
% nuance
Artificially shifting the AGB C yields to lower masses also improves agreement with the [C/Mg]-[Mg/Fe] trend, possibly indicating a discrepancy with stellar evolution predictions or our assumed Fe production rate.
\end{abstract}

@@

\begin{figure*}
    \centering
    \includegraphics{subgiants.pdf}
    \caption{The [C/Mg] ratio versus [Mg/H] (left) and [Mg/Fe] (right) for the \citet{jack}~sample of \apogee{} subgiants. {\bf Left:} High- and low-$\alpha$ stars are shown in blue and orange, respectively, using the separation defined in Eq.~\ref{eq:high_alpha}. The black points represent the median trend along the low-$\alpha$ sequence. {\bf Right:} Stars are colour-coded by their [Mg/H] abundance. The black points show the median [C/Mg]-[Mg/Fe] sequence for stars in the $-0.15 \leq [{\rm Mg/H}] \leq -0.05$ range. We use these median trends as our empirical benchmark in subsequent figures. } \label{fig:subgiants}
\end{figure*}

Despite being the second most abundant metal in the Universe~\citep[e.g.,][]{magg+22}, the nucleosynthetic origin of C is poorly understood. 
There is broad agreement that both massive and asymptotic giant branch (AGB) stars produce C \citep[e.g.,][]{jennifer19}, but their relative yields are disputed (see, e.g., the reviews by \citealt{romano22} and \citealt{RM21}). 
From a Galactic chemical evolution (GCE) perspective, some authors have argued that massive stars should dominate~\citep[e.g.,][]{prantzos+94, HEK00, romano+20, franchini+20, gustafsson22}, and others have argued that AGB stars should contribute a majority of C in the Universe~\citep[e.g.,][]{tinsley79, chiappini+03, mattsson10, KKU11, KKL20}. In addition, some studies like \citet{rybizki+17} find that AGB stars may produce about 20 to 50 per cent of solar C depending on the yield tables assumed.
In this paper, we aim to constrain C production by comparing abundance measurements in subgiant stars with multi-zone GCE models.%

Predictions of C yields from stellar evolution and supernova (SN) models are uncertain for numerous reasons (see, e.g., the reviews by \citealt{romano+10, KL14}).
Some prominent unknowns include mass loss rates \citep{sukhbold+16, beasor+2020}, rotational mixing \citep{frischknecht+16, LC18}, nuclear reaction rates \citep{herwig+05, herwig+austin2004}, and convection \citep{chieffi2001, ventura+13, issa+25}.
Recent work additionally suggests that small changes in a star's initial mass of 
${\sim}0.01\,\Mo$  have significant effects on the evolution and explosion outcomes of massive stars \citep{bruenn+2023, vartanyan_burrows2023}. Most yield tables have far courser initial-mass resolution.  Furthermore, many stars are found in binary systems \citep{sana+12}, the effects of which are almost entirely
unexplored in stellar nucleosynthesis models \citep{farmer+21}.
Robust predictions of metal yields from stellar evolution models therefore remain elusive.

Given the challenge of theoretical yield predictions, previous work has attempted to constrain chemical yields empirically. 
Even early chemical evolution work, such as \citet{garnett1997}, began to test stellar theory with observed abundance trends.
More recently, \citet{weinberg+19, weinberg+22} analysed the abundance ratio trends of high- and low-$\alpha$ sequence stars in APOGEE~\citep{apogee17} to infer relative yields of prompt and delayed enrichment sources and their metallicity dependence.
\citet{emily+19, emily+22, emily+24} applied the same methodology to GALAH~\citep{DeSilva2015, Martell2017}. In addition, models such as \citet{eitner+2020} and \citet{dubay+24} have used abundance trends of Fe and related elements to understand the physics and origin of \ia.
Most directly relevant to our analysis, \citet{james+24} applied Milky Way chemical evolution models to N and recommended a yield prescription that reproduces the observed relation between N and O abundances. We  extend their methodology to C in this paper.

% \citet{rodriguez+21, rodriguez+23} measured the mean Fe yield of core-collapse supernovae (CCSN) using the radioactive tails of their lightcurves.
% Their measurements, provide one of few empirical anchors on the absolute scale of yields \citep{david_fe}.

We use a sample of subgiant stars from APOGEE \citep{apogee17} as our primary observational constraint.
%This provide a large, homogeneous sample of well-measured C abundances unaffected by stellar evolution.
Subgiants' photospheric C abundance closely matches their birth composition \citep{gilroy89, korn+07, lind+08, souto+18, souto19}.
In other evolutionary phases, the measured abundances are known to be affected by internal processes (see discussion in Section~\ref{sec:data_selection} below).
Our sample should therefore be unaffected by uncertainties in inferring birth abundances.

In Section~\ref{sec:data_selection} we describe the selection for our subgiant observational sample.
In Section~\ref{sec:nucleosynthesis}, we discuss theoretical estimates of \agb\ and \cc\ C yields.
In Section~\ref{sec:vice}, we describe the details of our GCE models.
In Section~\ref{sec:results}, we present our model predictions and discuss their observationally distinguishable predictions. 
We then infer constraints on yield parameters in Section~\ref{sec:mcmc}.
In Section~\ref{sec:gas}, we compare our model to observations of halo stars and extragalactic \Hii{} regions from the literature. 
We summarize our conclusions in Section~\ref{sec:conclusions}.

@@

% Subgiants are a useful empirical benchmark because their photospheric abundances should reflect their birth mixture. 
We use subgiant stars as our empirical benchmark.
Subgiants have the advantage that their photospheric C and N abundances should closely approximate their birth mixture.
During main sequence evolution, metals can fall below the convective envelope (\textit{gravitational settling}; e.g., \citealt{turncotte+98}).
When stars evolve off the main sequence and become subgiants, these metals are reincorporated into the convective envelope \citep[]{gratton+00, souto19}. 
However, once a star becomes a red giant, \textit{first dredge up} pollutes its photosphere with CNO-processed material from the core \citep{iben67, KL14}.

We use the sample of \apogee\ DR17 subgiants from \citet{jack}. We select stars from the ASPCAP pipeline (revision 1) within the $\log g$-$T_\text{eff}$ polygon given by
 \begin{equation}
    \begin{cases} \label{eq:subgiant_selection}
        \log \text{g} \geq 3.5 \\
        \log \text{g} \leq 0.004\,T_{\rm eff} - 15.7 \\
        \log \text{g} \leq 0.0007\,T_{\rm eff} + 0.36 \\
        \log \text{g} \leq -0.0015\,T_{\rm eff} + 12.05 \\
        \log \text{g} \geq 0.0012\,T_{\rm eff} - 2.8. \\
    \end{cases}
\end{equation}
Following \citet{jack}, we also exclude stars with the flags:
        \verb|APOGEE_MIRCLUSTER_STAR|,
        \verb|APOGEE_EMISSION_STAR|,
        \verb|APOGEE_EMBEDDEDCLUSTER_STAR|,
        \verb|APOGEE2_YOUNG_CLUSTER|,
        \verb|APOGEE2_W345|,
        \verb|APOGEE2_EB|, and
        \verb|DUPLICATE|.
We additionally remove stars with no reported (or flagged unreliable) C, N, Mg, or Fe abundances. These cuts yield a sample containing \nsubgiants\ subgiants.

Fig.~\ref{fig:subgiants} shows the [C/Mg] abundance ratios of our subgiant sample as a function of [Mg/H] and [Mg/Fe].\footnotemark{} In this sample, [C/Mg] increases with [Mg/H], and [C/Mg] decreases with [Mg/Fe] at fixed [Mg/H].
Following \citet[eq.~1]{jack}, low-$\alpha$ stars are defined to have 
\begin{equation}\label{eq:high_alpha}
\begin{cases}
\text{[Mg/Fe]} <0.16-0.13\,\text{[Fe/H]}, & \text{[Fe/H]}<0\\
\text{[Mg/Fe]} <0.16, & \text{[Fe/H]}>0. \\
\end{cases}
\end{equation}
Compared to the low-$\alpha$ sequence in \caah, the high-$\alpha$ sequence has a systematically lower [C/Mg] value and steeper slope. This is consistent with the shape of the \caafe{} trend.
Note that the \caafe{} trend follows a two-component linear distribution, with a separation in [Mg/Fe] corresponding to the gap between high- and low-$\alpha$ stars. Additionally, the \caafe{} trend is approximately independent of metallicity. We aim to reproduce these trends in our GCE models.

\footnotetext{In this paper, we use the standard notation for chemical abundances. $[A/B] = \log_{10}\left(A/B\right) - \log_{10}\left(A_{\sun}/B_{\sun}\right)$, i.e. $[A/B]$ is the logarithm of the mass ratio between A and B, scaled such that $[A/B]=0$ for the Sun (see Table~\ref{tab:fiducial_mod} for solar scale.) 
}

While any abundance trend likely suffers from systematic biases, we find that these biases do not impact our qualitative conclusions. 
We have also consider subgiants and dwarfs with similar selection criteria from GALAH DR4 \citep{galah_dr4}, Gaia-ESO DR5.1 \citep{gaiaeso_dr5.1}, LAMOST \citep{lamost}, and mixing-corrected APOGEE giants \citep{vincenzo+21}. 
All of these samples show similar trends in \caah\ and \caafe{}, but with global systematic offsets of ${\sim} 0.1\,{\rm dex}$ and slight differences in the slope of each trend. Global offsets would affect our inferred total C trend with metallicity. However, given that the \caafe{} trend is similar among different surveys, we would recover similar results for the inferred relative contribution of AGB stars to C production (Section~\ref{sec:mcmc_results}). Our results should therefore be robust against systematic differences between these surveys.

\section{Carbon Nucleosynthesis}\label{sec:nucleosynthesis}
Here, we review the processes that produce C in both high- and low-mass stars. We define yields as used in this paper and discuss the physics, predictions, and uncertainties from stellar model predictions. Our yield assumptions for other elements are presented in Section~\ref{sec:vice}.

A yield quantifies the new production of a chemical species. For a given star, we define the {\it stellar yield of $X$} as the fraction of a star's initial mass synthesized into an element $X$ and released.  Specifically, we use {\it net} yields here, which subtracts the birth abundance of a chemical element from the ejected material. Net yields may become negative if more $X$ is transformed into other elements or locked inside the remnant than is produced. In contrast, a {\it gross} yield is the total mass of an element ejected over the star's lifetime and is always non-negative.
 We also describe yields by their population-averaged quantities, or {\it integrated yields}, which refer to the amount of newly produced $X$ per unit mass of star formation. These yields are calculated from the stellar mass-weighted average of net stellar yields over the initial mass function (IMF).

\subsection{Low and intermediate mass stars}\label{sec:agb}

\begin{figure*}
    \centering
 	    \includegraphics[scale=1]{agb_yields.pdf}
        \caption[]{The net fractional C yield from AGB stars as a function of initial stellar mass and metallicity. Each panel represents yields from one of four \agb\ studies from the literature: \fruity{}, \aton{}, \monash{}, \nugrid{} (see Section \ref{sec:agb}).  The black dotted line shows $\Ycagb=0$ for reference. 
        }

@@

Data and code used in this paper are happily available upon request.
